---
title: 'PS5: gapminder'
author: "Kristine"
date: "2023-02-16"
output: html_document
---
# 1 Load and check data
1. Load and check data:
```{r setup, message=FALSE}
library(tidyverse)
```
2. How many rows/columns do we have?
```{r, message=FALSE}
gm <- read_delim("~/info201/PS5/gapminder.csv")
gm
```
There are 13,055 rows and 25 columns.

3. Print a small sample of data. Does it look OK?
```{r, message=FALSE}
gm %>%
  sample_n(10)
```
Data looks good.

# 2 Descriptive statistics

1. How many countries are there in the dataset? Analyze all three: iso3, iso2 and name.
```{r}
gm %>%
  distinct(gm$name)%>% 
  count()
```
There are 250 under 'name'
```{r}
gm %>% 
  distinct(gm$iso2) %>% 
  count()
```
There are 249 under 'iso2'
```{r}
gm %>% 
  distinct(gm$iso3) %>% 
  count()
```
There are 253 under 'iso3'

2. If you did this correctly, you saw that there are more names than iso-2 codes, and there are
even more iso3 -codes. What is going on? Can you find it out?

```{r, message=FALSE}
gm %>% 
  group_by(iso2) %>% 
  summarize(num = n_distinct(name), na.rm = TRUE)
```
a. Find how many names are there for each iso-2 code. Are there any iso-2 codes that
correspond to more than one name? What are these countries?
```{r}
gm %>% 
  group_by(name) %>% 
  select(iso2, name) %>% 
  summarize(num = n_distinct(iso2)) %>% 
  arrange(desc(num))
```
There are 2 iso-2 codes that correspond to more than one name. The country is Afghanistan

b. Now repeat the same for name and iso3-code. Are there country names that have
more than one iso3-code? What are these countries?
```{r, message=FALSE}
gm %>% 
  group_by(iso3) %>% 
  select(iso3, name) %>% 
  summarize(num = n_distinct(iso3)) %>% 
  arrange(desc(num))
```
There are 4 countries that have more than one iso-3 code.

3. What is the minimum and maximum year in these data?
```{r, message=FALSE}
min_year <- min(gm$time)
max_year <- max(gm$time)

cat("The minimum year in the dataset is", min_year, "and the maximum year is", max_year, ".")
```


# 3 CO2 emissions

1. How many missing co2 emissions are there for each year? Analyze both missing CO2
and co2_PC. Which years have most missing data?
```{r}
missing_co2 <- gm %>%
  group_by(time) %>%
  summarize(num_missing_co2 = sum(is.na(CO2)),
            num_missing_co2_pc = sum(is.na(co2_PC)))
years_with_most_missing_co2 <- missing_co2 %>%
  arrange(desc(num_missing_co2)) %>%
  head(5)
years_with_most_missing_co2_pc <- missing_co2 %>%
  arrange(desc(num_missing_co2_pc)) %>%
  head(5)
```


2. Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few
more countries of your choice. Explain what do you see.
```{r}
countries <- c("United States", "China", "India", "Germany", "Japan")
co2_emissions <- gm %>%
  filter(region %in% countries)

co2_emissions$time <- as.numeric(co2_emissions$time)

ggplot(co2_emissions, aes(x = time, y = co2_emissions)) +
  geom_line(y = 1, x = 1)
  labs(title = "Total CO2 Emissions over Time",
       x = "Year", y = "Total CO2 Emissions (metric tons per capita)")
```
#I am confused with this question

3. Now let’s analyze the CO2 emissions per capita (co2_PC ). Make a similar plot of the
same countries. What does this figure suggest?
#I am struggling with this one as well.

4. Compute average CO2 emissions per capita across the continents (assume region is the
same as continent). Comment what do you see.
Note: just compute averages over countries and ignore the fact that countries are of different
size.
Hint: Americas 2016 should be 4.80.
```{r, message=FALSE}
avg_co2 <- gm %>% 
  select(region, co2_PC, time) %>% 
  filter(!is.na(region), time %in% c("1960", "2016")) %>% 
  group_by(region, time) %>% 
  summarize(avg_co2_PC = mean(co2_PC, na.rm = TRUE))

ggplot(avg_co2, aes(region, avg_co2_PC, fill=factor(time))) +
  geom_col(position = "dodge") +
  labs(title = "Average CO2 per Capita for Regions (Continents)",
       x = "Region (Continent Name)",
       y = "Average CO2 per Capita (Metric Ton per Capita)",
       fill = "Time (Year)")
```
5. Make a barplot where you show the previous results–average CO2 emissions per capita
across continents in 1960 and 2016.
```{r, message=FALSE}
avg_co2 <- gm %>% 
  select(region, co2_PC, time) %>% 
  filter(!is.na(region), time %in% c("1960", "2016")) %>% 
  group_by(region, time) %>% 
  summarize(avg_co2_PC = mean(co2_PC, na.rm = TRUE))

ggplot(avg_co2, aes(region, avg_co2_PC, fill=factor(time))) +
  geom_col(position = "dodge") +
  labs(title = "Average CO2 per Capita for Regions (Continents)",
       x = "Region (Continent Name)",
       y = "Average CO2 per Capita (Metric Ton per Capita)",
       fill = "Time (Year)")
```


6. Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per
capita) in 2019 for each continent? (Assume region is continent).

Three largest:
```{r, message = FALSE}
gm %>% 
  drop_na(region, time) %>% 
  filter(!is.na(co2_PC)) %>% 
  filter(time == "2016") %>% 
  group_by(region, name) %>% 
  summarize(avg_co2_pc = mean(co2_PC), groups = "drop") %>% 
  arrange(region, avg_co2_pc) %>% 
  group_by(region) %>% 
  tail(3) %>% 
  select(region, name, avg_co2_pc)
```
Three smallest:
```{r, message = FALSE}
gm %>% 
  drop_na(region, time) %>% 
  filter(!is.na(co2_PC)) %>% 
  filter(time == "2016") %>% 
  group_by(region, name) %>% 
  summarize(avg_co2_pc = mean(co2_PC), groups = "drop") %>% 
  arrange(region, avg_co2_pc) %>% 
  group_by(region) %>% 
  head(3) %>% 
  select(region, name, avg_co2_pc)
```

# 4 GDP per capita

1. Make a scatterplot of GDP per capita versus life expectancy by country, using data for
1960. Make the point size dependent on the country size, and color those according to the
continent. Feel free to adjust the plot in other ways to make it better.
Comment what do you see there.

```{r, warning=FALSE}
gdp_PC <- gm %>% 
  group_by(name) %>% 
  select(region, GDP_PC, lifeExpectancy, totalPopulation, time) %>% 
  filter(time == "1960")

ggplot(gdp_PC, aes(x = GDP_PC, y = lifeExpectancy, col = factor(region))) +
  geom_point(aes(size = totalPopulation / 1000000),
             alpha = 0.8) +
  labs(title = "Life Expectancy vs. GDP per Capita in 1960",
       x = "GDP per Capita", 
       y = "Life Expectancy (Years)",
       col = "Region (Continents)",
       size = "Total Population (Millions)")
```

I see a scatterplot trending 

2. Make a similar plot, but this time use 2019 data only.

```{r, warning=FALSE}
gdp_PC <- gm %>% 
  group_by(name) %>% 
  select(region, GDP_PC, lifeExpectancy, totalPopulation, time) %>% 
  filter(time == "2019")

ggplot(gdp_PC, aes(x = GDP_PC, y = lifeExpectancy, col = factor(region))) +
  geom_point(aes(size = totalPopulation / 1000000))
```

3. Compare these two plots and comment what do you see. How has world developed
through the last 60 years?
```{r}
gm %>% 
  filter(!is.na(GDP_PC), time %in% c(1960, 2019)) %>% 
  ggplot(aes(x = GDP_PC, fill = factor(time))) +
  geom_histogram(alpha = 0.5, position = "dodge", bins = 30) +
  scale_fill_manual(values = c("blue", "lightgreen"), labels =
c("1960", "2019"))+
  labs(x = "GDP per capita", y = "Count", title = "GDP per capita for 1960 and 2019") +
  theme_minimal()
```
4. Compute the average life expectancy for each continent in 1960 and 2019. Do the results
fit with what do you see on the figures?
Note: here as average I mean just average over countries, ignore the fact that countries are of
different size.
```{r}
gm %>% 
  group_by(time) %>% 
  select(co2_PC) %>% 
  filter(is.na(co2_PC)) %>% 
  count()
gm %>% 
  filter(time == "2016") %>% 
  group_by(region) %>% 
  summarize(co2_pc_avg = mean(co2_PC, na.rm = TRUE))
```

5. Compute the average LE growth from 1960-2019 across the continents. Show the results
in the order of growth. Explain what do you see.
Hint: these data (data in long form) is not the simplest to compute growth. But you may
want to check out the lag() function. And do not forget to group data by continent when
using lag(), otherwise your results will be messed up! See https://faculty.washington.
edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute.

```{r, message=FALSE}
avg_LE <- gm %>% 
  group_by(region) %>% 
  mutate(prev.val = lag(lifeExpectancy)) %>% 
  filter(row_number() %% 2 == 0) %>% 
  mutate(avg_growth = lifeExpectancy - prev.val) %>% 
  summarize(avg_growth = mean(avg_growth, na.rm = TRUE)) %>% 
  arrange(desc(avg_growth))

```

6. Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both
histograms on the same graph, see how well you can do it!

7. What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When
counting from top.)
Hint: check out the function rank()!
Hint2: 17 for 1960.
```{r}
ranks_per_year <- gm %>% 
  select(name, lifeExpectancy, time) %>% 
  filter(!is.na(lifeExpectancy), !is.na(name), time %in% c("1960", "2019")) %>% 
  arrange(time, desc(lifeExpectancy)) %>% 
  group_by(time) %>% 
  mutate(rankings = rank(desc(lifeExpectancy)))

ranks_per_year %>% 
  filter(name == "United States of America")
```

8. If you did this correctly, then you noticed that US ranking has been falling quite a
bit. But we also have more countries in 2019–what about the relative rank divided by the
corresponding number of countries that have LE data in the corresponding year?
Hint: 0.0904 for 1960.

```{r}
ranks_per_year %>% 
  select(name, time, rankings) %>% 
  group_by(time) %>% 
  mutate(total_n = max(rankings)) %>% 
  filter(name == "United States of America") %>% 
  summarize(rr_over_ncountries = rankings/total_n)
```


Finally tell us how many hours did you spend on this PS.
16 hours
